---
title: "Causal Inference with Continuous Exposures: A Tutorial with Application to ICU Data: Real Data Analysis"
author: "M Ehsan Karim"
output: 
  pdf_document:
    toc: false
    number_sections: true
    fig_caption: true
    citation_package: natbib
    keep_tex: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(nnet)
library(survey)
library(Publish)
library(tmle3)
library(tmle3shift)
library(sl3)
library(data.table)
library(future)
library(tableone)
library(MatchIt)
library(cobalt)
```

## Download and Prepare RHC Data

```{r data-download, cache=TRUE}
# Download data
ObsData <- read.csv("https://hbiostat.org/data/repo/rhc.csv", header = TRUE)

# Calculate Length of Stay
ObsData$Length.of.Stay <- ObsData$dschdte - ObsData$sadmdte
ObsData$Length.of.Stay[is.na(ObsData$Length.of.Stay)] <- 
  ObsData$dthdte[is.na(ObsData$Length.of.Stay)] - 
  ObsData$sadmdte[is.na(ObsData$Length.of.Stay)]

# Binary outcome
ObsData$death <- ifelse(ObsData$death == "Yes", 1, 0)

# Remove unwanted outcome variables
ObsData <- dplyr::select(ObsData, !c(dthdte, 
                                     lstctdte, 
                                     dschdte, 
                                     t3d30, 
                                     dth30, 
                                     surv2md1))

# Remove problematic variables
ObsData <- dplyr::select(ObsData, !c(sadmdte, 
                                     ptid, 
                                     X, 
                                     adld3p, 
                                     urin1, 
                                     cat2))

# Convert categorical variables
factors <- c("cat1", "ca", "death", "cardiohx", "chfhx", "dementhx", "psychhx", 
             "chrpulhx", "renalhx", "liverhx", "gibledhx", "malighx", "immunhx", 
             "transhx", "amihx", "sex", "dnr1", "ninsclas", "resp", "card", "neuro", 
             "gastr", "renal", "meta", "hema", "seps", "trauma", "ortho", "race", 
             "income")
ObsData[factors] <- lapply(ObsData[factors], as.factor)

# Recode RHC use
ObsData$RHC.use <- ifelse(ObsData$swang1 == "RHC", 1, 0)
ObsData <- dplyr::select(ObsData, -swang1)

# Recode and factor levels
ObsData$age <- cut(ObsData$age, breaks=c(-Inf, 50, 60, 70, 80, Inf), right=FALSE)
ObsData$race <- factor(ObsData$race, levels=c("white", "black", "other"))
ObsData$sex <- relevel(as.factor(ObsData$sex), ref = "Male")
ObsData$cat1 <- factor(ObsData$cat1, levels = unique(ObsData$cat1))
levels(ObsData$cat1) <- c("ARF", "CHF", "Other", "Other", "Other", "Other", "Other", "MOSF", "MOSF")
ObsData$ca <- factor(ObsData$ca, levels = c("No", "Yes"), labels = c("None", "Metastatic"))

# Rename variables
names(ObsData) <- c("Disease.category", "Cancer", "Death", "Cardiovascular", "Congestive.HF", 
                    "Dementia", "Psychiatric", "Pulmonary", "Renal", "Hepatic", "GI.Bleed", 
                    "Tumor", "Immunosupperssion", "Transfer.hx", "MI", "age", "sex", "edu", 
                    "DASIndex", "APACHE.score", "Glasgow.Coma.Score", "blood.pressure", "WBC", 
                    "Heart.rate", "Respiratory.rate", "Temperature", "PaO2vs.FIO2", "Albumin", 
                    "Hematocrit", "Bilirubin", "Creatinine", "Sodium", "Potassium", "PaCo2", 
                    "PH", "Weight", "DNR.status", "Medical.insurance", "Respiratory.Diag", 
                    "Cardiovascular.Diag", "Neurological.Diag", "Gastrointestinal.Diag", 
                    "Renal.Diag", "Metabolic.Diag", "Hematologic.Diag", "Sepsis.Diag", 
                    "Trauma.Diag", "Orthopedic.Diag", "race", "income", "Length.of.Stay", 
                    "RHC.use")

str(ObsData)
# Save cleaned dataset
saveRDS(ObsData, file = "rhcAnalytic.RDS")
```

## Load RHC Data

```{r load, cache=TRUE}
rhc <- readRDS("rhcAnalytic.RDS")

# Define comprehensive confounder set
covariates <- c(
  "age", "sex", "Cardiovascular", "Pulmonary", 
  "Renal", "Congestive.HF", "Cancer",
  "APACHE.score", "DASIndex", "Albumin", 
  "Creatinine", "Sodium", "Heart.rate", "WBC",
  "DNR.status", "Transfer.hx"
)

# Prepare variables
rhc <- rhc %>%
  mutate(
    A = blood.pressure,
    Y = as.numeric(as.character(Death))
  ) %>%
  filter(!is.na(A), !is.na(Y)) %>%
  filter(complete.cases(select(., all_of(covariates))))
```

## Summarize the Exposure

```{r plot-A-distribution, message=FALSE, warning=FALSE, cache=TRUE}
ggplot(rhc, aes(x = A)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "steelblue", alpha = 0.6) +
  geom_density(color = "black", linewidth = 1) +
  labs(
    title = "Distribution of Blood Pressure (Exposure A)",
    x = "Blood Pressure",
    y = "Density"
  ) +
  theme_minimal(base_size = 13)
```

## Method 1: IPW with Normal Exposure Model

```{r ipw-normal, cache=TRUE}
# Numerator: marginal exposure model
mod_num <- lm(A ~ 1, data = rhc)
mu_num <- predict(mod_num)
sd_num <- sd(residuals(mod_num))

# Denominator: exposure conditional on confounders
form_denom <- as.formula(paste("A ~", 
                               paste(covariates, collapse = " + ")))
mod_denom <- lm(form_denom, data = rhc)
mu_denom <- predict(mod_denom)
sd_denom <- sd(residuals(mod_denom))

# Densities and weights
f_num <- dnorm(rhc$A, mean = mu_num, sd = sd_num)
f_denom <- dnorm(rhc$A, mean = mu_denom, sd = sd_denom)
rhc$sw_normal <- f_num / f_denom

# Weighted model
mod_w_normal <- glm(Y ~ A, family = binomial(), 
                    data = rhc, weights = sw_normal)
publish(mod_w_normal)
```

## Method 2: IPW with Quantile Binning

```{r ipw-qbin, cache=TRUE}
# Create quantile bins
rhc$qbin <- cut(rhc$A, breaks = quantile(rhc$A, 
                                         probs = seq(0, 1, 0.1), 
                                         na.rm = TRUE),
                include.lowest = TRUE, labels = FALSE)

# Multinomial model for conditional bin probability
form_q <- as.formula(paste("factor(qbin) ~", 
                           paste(covariates, collapse = " + ")))
mod_q <- multinom(form_q, data = rhc, trace = FALSE)
p_denom <- predict(mod_q, type = "probs")
row_idx <- cbind(1:nrow(rhc), rhc$qbin)
p_denom_val <- p_denom[row_idx]
p_num_val <- 1 / 10
rhc$sw_qbin <- p_num_val / p_denom_val

# Weighted model
mod_w_qbin <- glm(Y ~ A, family = binomial(), 
                  data = rhc, weights = sw_qbin)
publish(mod_w_qbin)
```

## Method 3: TMLE with Shift Intervention

```{r tmle-shift, cache=FALSE}
# TMLE setup
node_list <- list(W = covariates, A = "A", Y = "Y")
glm_learner <- make_learner(Lrnr_glm)
learner_list <- list(Y = glm_learner, A = glm_learner)

tmle_spec <- tmle_shift(
  delta = 0.1,
  shift_fxn = function(tmle_task, delta, ...) {
    a <- tmle_task$get_tmle_node("A")
    a + delta
  },
  max_shift = 1
)

# Run TMLE
future::plan(future::sequential)
tmle_fit <- tmle3(
  tmle_spec,
  data = as.data.table(rhc),
  node_list = node_list,
  learner_list = learner_list
)

# Extract estimate
psi <- tmle_fit$estimates[[1]]$psi
IC <- tmle_fit$estimates[[1]]$IC
se <- sd(IC) / sqrt(length(IC))
ci <- psi + c(-1.96, 1.96) * se

# Output
data.frame(
  logOR = psi,
  OR = exp(psi),
  se = se,
  lower = exp(ci[1]),
  upper = exp(ci[2])
)
```


## Results

```{r forest-plot-or, message=FALSE, warning=FALSE, cache=TRUE}
# Manually collect results into a data frame
estimates <- data.frame(
  Method = c("IPW (Normal)", "IPW (Quantile Binning)", "TMLE (Shift 0.1)"),
  logOR = c(
    coef(mod_w_normal)["A"],
    coef(mod_w_qbin)["A"],
    psi
  ),
  SE = c(
    sqrt(vcov(mod_w_normal)["A", "A"]),
    sqrt(vcov(mod_w_qbin)["A", "A"]),
    sd(IC) / sqrt(length(IC))
  )
)

# Compute OR and 95% CI
estimates <- estimates %>%
  mutate(
    OR = exp(logOR),
    lower = exp(logOR - 1.96 * SE),
    upper = exp(logOR + 1.96 * SE)
  )

# Plot
ggplot(estimates, aes(x = Method, y = OR)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  coord_flip() +
  labs(
    title = "Estimated Odds Ratio for Effect of Blood Pressure on Death",
    y = "Odds Ratio (OR)",
    x = NULL
  ) +
  theme_minimal(base_size = 13)
```

```{r results-table, message=FALSE, warning=FALSE, cache=TRUE}
library(knitr)
library(kableExtra)

# Prepare table of OR estimates
results_table <- estimates %>%
  transmute(
    Method,
    OR = sprintf("%.3f", OR),
    `95% CI` = sprintf("%.3f – %.3f", lower, upper),
    SE = sprintf("%.3f", SE)
  )

# Show as styled table
kable(results_table, format = "latex", booktabs = TRUE, caption = "Estimated Odds Ratios and 95% Confidence Intervals")
```

## Save plot 

```{r save-plot, message=FALSE, warning=FALSE, cache=TRUE}
# Load required libraries
library(ggplot2)
library(patchwork)

# Plot 1: Distribution of the exposure A (Blood Pressure)
p1 <- ggplot(rhc, aes(x = A)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "grey70", color = "black") +
  geom_density(color = "black", linewidth = 1) +
  labs(
    title = "Distribution of Exposure: Blood Pressure",
    x = "Mean Arterial Blood Pressure (mm Hg)",
    y = "Density"
  ) +
  theme_minimal(base_size = 12)

# Plot 2: Forest plot of estimated odds ratios for death
p2 <- ggplot(estimates, aes(x = Method, y = OR)) +
  geom_point(size = 3, color = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.15, color = "black") +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey50") +
  coord_flip() +
  labs(
    title = "Estimated Association with In-Hospital Death",
    y = "Odds Ratio (per 1 mm Hg increase in Blood Pressure)",
    x = NULL
  ) +
  theme_minimal(base_size = 12)

# Combine plots side by side
combined_plot <- p1 + p2 + plot_layout(ncol = 2)

# Display
print(combined_plot)

# Optional: Save the plot
ggsave("rhc_bw_combined_plot.png", combined_plot, width = 12, height = 5, dpi = 600)
```

## Diagnostic Checks

To assess the reliability of causal estimates, we conduct several diagnostic checks related to the validity of assumptions and model performance. These include checking weight stability, covariate balance, and the positivity assumption.

### Weight Diagnostics

```{r weight-diagnostics, message=FALSE, warning=FALSE, cache=TRUE}
# Histograms of weights
summary(rhc$sw_normal)
summary(rhc$sw_qbin)
par(mfrow = c(1, 2))
hist(rhc$sw_normal, main = "IPW Normal Weights", xlab = "Weight", col = "grey80", border = "white")
hist(rhc$sw_qbin, main = "IPW Quantile Binning Weights", xlab = "Weight", col = "grey80", border = "white")
```

Extreme or highly variable weights can indicate misspecification in the exposure model or violations of the positivity assumption.

### Covariate Balance with IPW

```{r covariate-balance, message=FALSE, warning=FALSE, cache=TRUE}
# Covariate balance before and after IPW using quantile binning
bal <- bal.tab(
  as.formula(paste("A ~", paste(covariates, collapse = " + "))),
  data = rhc,
  weights = rhc$sw_qbin,
  method = "weighting"
)
bal
love.plot(bal, abs = TRUE, thresholds = c(m = 0.1))
```

The love plot shows performance before and after weighting. Good balance is typically indicated by values below 0.1.

### Positivity Check

```{r positivity-check, message=FALSE, warning=FALSE, cache=TRUE}
# Exposure distribution by confounder category
ggplot(rhc, aes(x = A, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(title = "Blood Pressure by Sex (as a Proxy for Positivity)", x = "Blood Pressure", y = "Density") +
  theme_minimal()
```

This density plot illustrates whether the exposure is sufficiently variable across levels of key confounders. Lack of overlap may indicate a positivity violation.
